% Part: sets-functions-relations
% Chapter: sets
% Section: pairs-and-products
%
%\documentclass[../../../include/open-logic-section]{subfiles}
%
\begin{document}
%
\olfileid{sfr}{set}{pai}
\olsection{Pairs, Tuples, Cartesian Products}\ollabel{pai}
\begin{explain}
	As we remarked, it follows from Extensionality that sets have no order to their elements. So if we want to represent order, we use \emph{ordered pairs} $\tuple{x, y}$. In an unordered pair $\{x, y\}$, the order does not matter: $\{x, y\} = \{y, x\}$. In an ordered pair, it does: if $x \neq y$, then $\tuple{x, y} \neq \tuple{y, x}$.

How should we think about ordered pairs in set theory? Crucially, we want to preserve the idea that ordered pairs are identical iff they share the same first element and share the same second element, i.e.:
	$$\tuple{a, b}= \tuple{c, d}\text{ iff both }a = c \text{ and }b=d $$
Normally, this is implemented using the Wiener--Kuratowski definition (I leave it as an exercise to prove that this definition preserves the biconditional).
\end{explain}
\begin{defn}[Ordered pair]\ollabel{wienerkuratowski}
	$\tuple{a, b} = \{\{a\}, \{a, b\}\}$.
\end{defn}
\begin{prob}
	Using \olref[sfr][set][pai]{wienerkuratowski}, prove that $\tuple{a, b}= \tuple{c, d}$ iff both $a = c$ and $b=d$.
\end{prob}
\begin{explain}
	Having fixed a definition of an ordered pair, we can use it to define further sets. For example, sometimes we also want ordered sequences of more than two objects,
	e.g., \emph{triples} $\tuple{x, y, z}$, \emph{quadruples} $\tuple{x,
		y, z, u}$, and so on.  We can think of triples as special
	ordered pairs, where the first element is itself an ordered pair:
	$\tuple{x, y, z}$ is $\tuple{\tuple{x, y},z}$. The same is
	true for quadruples: $\tuple{x,y,z,u}$ is 
	$\tuple{\tuple{\tuple{x,y},z},u}$, and so on. In general, we talk of
	\emph{ordered $n$-tuples} $\tuple{x_1, \dots, x_n}$.

	Certain sets of ordered pairs, or other ordered $n$-tuples, will be useful.
\end{explain}
	\begin{defn}[Cartesian product]
		Given sets $A$ and $B$, their \emph{Cartesian product} $A \times B = \Setabs{\tuple{x, y}}{x \in A \text{ and } y \in B}$.
	\end{defn}
	
	\begin{ex}
		If $A = \{0, 1\}$, and $B = \{1, c, d\}$, then their product is
		\[
			A\times B = \{ \tuple{0, 1}, \tuple{0, c}, \tuple{0, d},
		\tuple{1, 1}, \tuple{1, c}, \tuple{1, d} \}.
		\]
	\end{ex}
	
	\begin{ex}
		If $A$ is a set, the product of $A$ with itself, $A \times A$, is also
		written~$A^2$. It is the set of \emph{all} pairs $\tuple{x, y}$ with
		$x, y \in A$, i.e.\ $A \times A = \Setabs{\tuple{x,y}}{x, y \in A}$. The set of all triples $\tuple{x, y, z}$ is $A^3$, and
		so on. We can give a recursive definition:
		\begin{align*}
			A^1 & = A\\
			A^{n+1} & = A^n \times A
		\end{align*}
	\end{ex}
	
	\begin{prob}
		List all !!{element}s of $\{1, 2, 3\}^3$.
	\end{prob}
	
	\begin{prop}\ollabel{cardnmprod}
		If $A$ has $n$ !!{element}s and $B$ has $m$ !!{element}s, then $A
		\times B$ has $n\cdot m$ elements.
	\end{prop}
	\begin{proof}
		For every $x \in A$ there are $m$ !!{element}s of the form $\tuple{x, y} \in A \times B$. Let $B_x = \Setabs{\tuple{x, y}}{y \in B}$. Whenever $x_1 \neq x_2$, we have $\tuple{x_1, y} \neq
		\tuple{x_2, y}$, and so $B_{x_1} \cap B _{x_2} = \emptyset$. But if $A = \{x_1,
		\dots, x_n\}$, then $A \times B = B_{x_1} \cup \dots \cup B_{x_n}$, and so has
		$n\cdot m$ !!{element}s.
		
		To visualize this, arrange the !!{element}s of~$A \times B$ in a grid:
		\[
		\begin{array}{rcccc}
		B_{x_1} = & \{\tuple{x_1, y_1}, & \tuple{x_1, y_2}, & \dots & \tuple{x_1, y_m}\}\\
		B_{x_2} = & \{\tuple{x_2, y_1}, & \tuple{x_2, y_2}, & \dots & \tuple{x_2, y_m}\}\\
		\vdots & & \vdots\\
		B_{x_n} = & \{\tuple{x_n, y_1}, & \tuple{x_n, y_2}, & \dots & \tuple{x_n, y_m}\}
		\end{array}
		\]
		Since the $x_i$ are all different, and the $y_j$ are all different, no
		two of the pairs in this grid are the same, and there are $n\cdot m$
		of them.
	\end{proof}
	
	\begin{prob}
		Show, by induction on~$k$, that for all $k \ge 1$, if $X$ has $n$
		!!{element}s, then $X^k$ has $n^k$ !!{element}s.
	\end{prob}
	
%	\begin{ex}
%		If $X$ is a set, a \emph{word} over~$X$ is any sequence of
%		!!{element}s of~$X$. A sequence can be thought of as an $n$-tuple of
%		!!{element}s of~$X$. For instance, if $X = \{a, b, c\}$, then the
%		sequence ``$bac$'' can be thought of as the triple~$\tuple{b, a, c}$.
%		Words, i.e., sequences of symbols, are of crucial importance in
%		computer science, of course. By convention, we count !!{element}s of
%		$X$ as sequences of length~1, and $\emptyset$ as the sequence of
%		length~0. The set of \emph{all} words over~$X$ then is
%		\[
%		X^* = \{\emptyset\} \cup X \cup X^2 \cup X^3 \cup \dots
%		\]
%	\end{ex}

\end{document}