\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}
		\olfileid{sfr}{card-arithmetic}{card-opps}
\olsection{Defining the basic operations}\ollabel{card-opps}
Since we do not need to keep track of order, cardinal arithmetic is rather easier to define than ordinal arithmetic. I will define addition, multiplication, and exponentiation simultaneously. 
\begin{defn}When $\cardfont{a}$ and $\cardfont{b}$ are cardinals:
	\begin{align*}
	\cardfont{a} \cardplus \cardfont{b} &:= \cardof{\cardfont{a} \disjointsum \cardfont{b}}\\
	\cardfont{a} \cardtimes \cardfont{b} &:= \cardof{\cardfont{a} \times \cardfont{b}}\\
	\cardexpo{\cardfont{a}}{\cardfont{b}} &:= \cardof{\funfromto{\cardfont{b}}{\cardfont{a}}}
	\end{align*}
	where $\funfromto{X}{Y} = \Setabs{f}{f\text{ is a function }X \to Y}$. (It is easy to show that $\funfromto{X}{Y}$ exists for any sets $X$ and $Y$; I leave this as an exercise.) 
\end{defn}
\begin{prob}
	Prove in $\Zminus$ that $\funfromto{X}{Y}$ exists for any sets $X$ and $Y$. Working in $\ZF$, compute $\setrank{\funfromto{X}{Y}}$ from $\setrank{X}$ and $\setrank{Y}$, in the manner of \olref[sfr][ord-arithmetic][ord-addition]{rankcomputation}. 
\end{prob}\noindent
It might help to explain this definition. Concerning addition: this uses the notion of disjoint sum, $\disjointsum$, as defined in \olref[sfr][ord-arithmetic][ord-addition]{defdissum}; and it is easy to see that this definition gives the right verdict for finite cases. Concerning multiplication: \olref[sfr][set][pai]{cardnmprod} tells us that if $A$ has $n$ members and $B$ has $m$ members then $A \times B$ has $n \cdot m$ members, so our definition simply generalises the idea to transfinite multiplication. Exponentiation is similar: we are simply generalising the thought from the finite to the transfinite. Indeed, in certain ways, transfinite cardinal arithmetic looks much more like ``ordinary'' arithmetic than does transfinite ordinal arithmetic:
\begin{prop}\ollabel{cardplustimescommute} $\cardplus$ and $\cardtimes$ are commutative and associative. 
\end{prop}
\begin{proof}
	For commutativity, by \olref[sfr][cardinals][card-cardsasords]{lem:CardinalsBehaveRight} it suffices to observe that $\cardeq{(\cardfont{a} \disjointsum \cardfont{b})}{(\cardfont{b} \disjointsum \cardfont{a})}$ and $\cardeq{(\cardfont{a} \times \cardfont{b})}{(\cardfont{b} \times \cardfont{a})}$. I leave associativity to the reader.
\end{proof}
\begin{prob}
	Prove that $\cardplus$ and $\cardtimes$ are associative.
\end{prob}
\begin{prop}
	$A$ is infinite iff $\cardof{A} \cardplus 1 = 1 \cardplus \cardof{A} = \cardof{A}$
\end{prop}
\begin{proof}
	As in \olref[sfr][card-arithmetic][card-classing]{generalinfinitycharacter}, from \olref[sfr][ord-arithmetic][ord-addition]{ordinfinitycharacter} and \olref[sfr][cardinals][card-cardsasords]{lem:CardinalsBehaveRight}. 
\end{proof}\noindent
This explains why we need to use different symbols for ordinal versus cardinal addition/multiplication: these are genuinely \emph{different} operations. This next pair of results shows that ordinal versus cardinal exponentiation are also different operations. (Recall that  \olref[sfr][z][infinity-again]{defnomega} entails that $2 = \{0, 1\}$):
\begin{lem}\ollabel{lem:SizePowerset2Exp}$\cardof{\Pow{A}} = \cardexpo{2}{\cardof{A}}$, for any $A$
\end{lem}
\begin{proof}
	For each subset $B \subseteq A$, let $\chi_B \in \funfromto{A}{2}$ be given by:
	\begin{align*}
		\chi_{B}(x) &:=
		\begin{cases}
			1 & \text{if }x\in B\\
			0 & \text{otherwise}
		\end{cases}
	\end{align*}
	Now let $f(B) = \chi_B$; this defines !!a{bijection} $f \colon \Pow{A} \to \funfromto{A}{2}$. So $\cardeq{\Pow{A}}{\funfromto{A}{2}}$. Hence $\cardeq{\Pow{A}}{\funfromto{\cardof{A}}{2}}$, so that $\cardof{\Pow{A}} = \cardof{\funfromto{\cardof{A}}{2}} = 2^{\cardof{A}}$.
\end{proof}\noindent
This snappy proof essentially subsumes the discussion of \olref[sfr][siz][red]{red}. There, we showed how to ``reduce'' the uncountability of $\Pow{\omega}$ to the uncountability of the set of infinite binary strings, $\Bin^\omega$. In effect, $\Bin^{\omega}$ is just $\funfromto{\omega}{2}$; and the preceding proof showed that the reasoning we went through in \olref[sfr][siz][red]{red} will go through using any set $A$ in place of $\omega$. The result also yields a quick fact about cardinal exponentiation:
\begin{cor}\ollabel{cantorcor}$\cardfont{a} < \cardexpo{2}{\cardfont{a}}$ for any cardinal $\cardfont{a}$.
\end{cor}
\begin{proof}
	From Cantor's Theorem (\olref[sfr][siz][car]{thm:cantor}) and \olref{lem:SizePowerset2Exp}.
\end{proof}\noindent
So $\omega < \cardexpo{2}{\omega}$. But note: this is a result about \emph{cardinal} exponentiation. It should be contrasted with \emph{ordinal} exponentation, since in the latter case $\omega = \ordexpo{2}{\omega}$ (see 
\olref[sfr][ord-arithmetic][ord-expo]{ord-expo}).

Whilst we are on the topic of cardinal exponentiation, we can also be a bit more precise about the ``way'' in which $\Real$ is !!{nonenumerable}. 
\begin{thm}\ollabel{continuumis2aleph0}
	$\cardof{\Real} = \cardexpo{2}{\omega}$
\end{thm}
\begin{proof}[Proof skeleton] There are plenty of ways to prove this. The most straightforward is to argue that $\cardle{\Pow{\omega}}{\Real}$ and $\cardle{\Real}{\Pow{\omega}}$, and then use Schr\"oder--Bernstein to infer that $\cardeq{\Real}{\Pow{\omega}}$, and  \olref[sfr][card-arithmetic][card-opps]{lem:SizePowerset2Exp} to infer that $\cardof{\Real} = \cardexpo{2}{\omega}$. I leave it as an (illuminating) exercise for the reader to define injections $f \colon \Pow{\omega} \to \Real$ and $g \colon \Real \to \Pow{\omega}$.
\end{proof}
\begin{prob}
	Complete the proof of \olref[sfr][card-arithmetic][card-opps]{continuumis2aleph0}, by showing that $\cardle{\Pow{\omega}}{\Real}$ and $\cardle{\Real}{\Pow{\omega}}$.
\end{prob}


\end{document}