% Part: sets-functions-relations
% Chapter: relations
% Section: orders
%
\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{sfr}{rel}{ord}
\olsection{Orders}

\begin{explain}
Many of our comparisons involve describing some objects as being ``less than'', ``equal to'', or ``greater than'' other objects, in a certain respect. These involve \emph{order} relations. But there are different kinds of order relations. For instance, some require that any two objects be comparable, others don't. Some include identity (like~$\le$) and some exclude it (like~$<$). It will help us to have a taxonomy here.
\end{explain}

\begin{defn}[Preorder]
A relation which is both reflexive and transitive is called a
\emph{preorder.}  
\end{defn}

\begin{defn}[Partial order]
A preorder which is also anti-symmetric is called a
\emph{partial order}.
\end{defn}

\begin{defn}[Linear order]
A partial order which is also connected is called a
\emph{total order} or \emph{linear order.}
\end{defn}\noindent
Every linear order is also a partial order, and every partial order is
also a preorder, but the converses don't hold. 


\begin{ex}
The universal relation
on~$X$ is a preorder, since it is reflexive and transitive. But, if
$X$ has more than one !!{element}, the universal relation is not
anti-symmetric, and so not a partial order. %For a somewhat less silly example, consider the \emph{no longer than} relation $\preccurlyeq$ on~$\Bin^*$: $x \preccurlyeq y$ iff $\len{x} \le \len{y}$. This is a preorder (reflexive and transitive), and even connected, but not a partial order, since it is not anti-symmetric. For instance, $01 \preccurlyeq 10$ and $10 \preccurlyeq 01$, but $01 \neq 10$.
\end{ex}
\begin{ex}
An important partial order is the relation $\subseteq$ on a set of sets. This is not in general a linear order, since if $a \neq b$ and we consider $\Pow{\{a, b\}} = \{\emptyset, \{a\}, \{b\}, \{a,b\}\}$, we see that $\{a\} \nsubseteq \{b\}$ and $\{a\} \neq \{b\}$ and $\{b\} \nsubseteq \{a\}$.
\end{ex}
\begin{ex}
The relation of \emph{divisibility without remainder} also gives us a partial order which is not a linear order. (This is usually written
``$n\mid m$'', and we define it thus on a set $X$: $n \mid m$ iff $(\exists k \in X)m = kn$.) On $\Nat$, this is a partial order, but not a linear order: for instance, $2\nmid3$ and also
$3\nmid2$. Considered as a relation on $\Int$, divisibility is only a
preorder, since it is not anti-symmetric: $1\mid-1$ and $-1\mid1$ but
$1\neq-1$. 
\end{ex}
\begin{defn}[Irreflexivity]
A relation $R$ on $X$ is called \emph{irreflexive} if, for all $x \in
X$, $ \lnot Rxx$. 
\end{defn}

\begin{defn}[Asymmetry]
A relation $R$ on $X$ is called \emph{asymmetric} if for no pair $x,y\in
X$ we have $Rxy$ and $Ryx$. 
\end{defn}

\begin{defn}[Strict order]
A \emph{strict order} is a relation which is irreflexive, asymmetric,
and transitive.
\end{defn}

\begin{defn}[Total order]\ollabel{defstrictlinearorder}
A strict order which is also connected is called a
\emph{total order}. 
This is also sometimes called a \emph{strict linear order}. 
\end{defn}\noindent
Any strict order $R$ on~$A$ can be turned into a partial order
by adding the diagonal $\Id{A}$, i.e., adding all the pairs~$\tuple{x,
  x}$.  (This is called the \emph{reflexive closure} of~$R$.)
Conversely, starting from a partial order, one can get a strict
order by removing~$\Id{A}$. These next two results make this precise.
\begin{prop}\ollabel{stricttopartial}
	If $R$ is a strict order on~$X$, then $R^+ = R
    \cup \Id{X}$ is a partial order. Moreover, if $R$ is total, then $R^+$ is a linear order.
\end{prop}
\begin{proof}
    	Suppose $R$ is a strict order, i.e., $R \subseteq
      X^2$ and $R$ is irreflexive, asymmetric, and transitive. Let $R^+
      = R \cup \Id{X}$. We have to show that $R^+$ is reflexive,
      antisymmetric, and transitive.

      $R^+$ is clearly reflexive, since $\tuple{x,
        x} \in \Id{X} \subseteq R^+$ for all $x \in X$. 

      To show $R^+$ is antisymmetric, suppose for reductio that $R^+xy$ and $R^+yx$ but $x \neq y$. Since $\tuple{x,y} \in R \cup \Id{X}$, but $\tuple{x, y} \notin
      \Id{X}$, we must have $\tuple{x, y} \in R$, i.e.,
      $Rxy$. Exactly similarly,~$Ryx$. But this contradicts the
      assumption that $R$ is asymmetric.

      To establish transitivity, suppose that $R^+xy$ and $R^+yz$. If both $\tuple{x, y} \in
      R$ and $\tuple{y,z} \in R$, then $\tuple{x, z} \in R$
      since $R$~is transitive. Otherwise, either $\tuple{x, y} \in
      \Id{X}$, i.e., $x = y$, or $\tuple{y, z} \in \Id{X}$, i.e., $y =
      z$. In the first case, we have that $R^+yz$ by assumption, $x =
      y$, hence $R^+xz$. Similarly in the second case. In either case,
      $R^+xz$, thus, $R^+$ is also transitive.

      Concerning the ``Moreover'' clause, suppose $R$ is total, i.e.\ that $R$ is connected. So for all $x \neq y$, either $Rxy$ or~$Ryx$, i.e., either $\tuple{x, y} \in R$ or $\tuple{y, x} \in
      R$. Since $R \subseteq R^+$, this remains true of $R^+$, so $R^+$ is connected as well.
  \end{proof}
 \begin{prop}\ollabel{partialtostrict}
  	If $R$ is a partial order on~$X$, then $R^- = R
  	\setminus \Id{X}$ is a strict order. Moreover, if $R$ is linear, then $R^-$ is total.
  	\end{prop}
  \begin{proof}
  	This is left as an exercise.
\end{proof}
\begin{ex}
	$\le$ is the linear order corresponding to the total 
	order~$<$. $\subseteq$ is the partial order corresponding to
	the strict order~$\subsetneq$.
\end{ex}
\noindent
We will close with a simple result which establishes that total orders satisfy an Extensionality-like property:
\begin{prop}\ollabel{propextensionalityfortotalorders}
	If $<$ totally orders $A$, then: 
	$$(\forall a, b \in A)((\forall x \in A)(x < a \liff x < b) \lonlyif a = b)$$
\end{prop}
\begin{proof}
	Suppose $(\forall x \in A)(x < a \liff x < b)$. If $a < b$, then $a < a$, contradicting the fact that $<$ is irreflexive; so $a \nless b$. Exactly similarly, $b \nless a$. So $a = b$, as $<$ is connected.
\end{proof}\noindent
\begin{prob}
  Complete the proof of \olref[sfr][rel][ord]{partialtostrict}. 
\end{prob}


\end{document}