\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}
		\olfileid{sfr}{ordinals}{vn}	
\olsection{Von Neumann's construction of the ordinals}\ollabel{vn}
\olref[sfr][ordinals][ord-iso]{thm:woalwayscomparable} gives rise to a thought. We could introduce certain objects, called \emph{order types}, to go proxy for the well-orderings. Writing $\ordtype{A, <}$ for the order type of the well-ordering $\tuple{A, <}$, we would hope to secure the following two principles:
\begin{align*}
	\ordtype{A, <} = \ordtype{B, \lessdot}&\text{ iff }\ordeq{\tuple{A, <}}{\tuple{B, \lessdot}}\\
	\ordtype{A, <} < \ordtype{B, \lessdot}&\text{ iff }\ordeq{\tuple{A, <}}{\tuple{B_b, \lessdot_b}}\text{ for some }b \in B
\end{align*}
Moreover, we might hope to introduce order-types \emph{as certain sets}, just as we can introduce the natural numbers as certain sets. 

The most common way to do this---and the approach we will follow---is to define these order-types via certain \emph{canonical} well-ordered sets. These canonical sets were first introduced by von Neumann:
\begin{defn}
	The set $A$ is \emph{transitive} {iff} $(\forall x \in A)x \subseteq A$. Then $A$ is an \emph{ordinal} {iff} $A$ is transitive and well-ordered by $\in$.
\end{defn}\noindent
In what follows, I will use Greek letters for ordinals. It follows immediately from the definition that, if $\alpha$ is an ordinal, then $\tuple{\alpha, \in_\alpha}$ is a well-ordering, where $\in_\alpha = \Setabs{\tuple{x, y} \in \alpha^2}{x \in y}$. So, abusing notation a little, we can just say that $\alpha$ \emph{itself} is a well-ordering. 

Here are our first few ordinals:
	$$\emptyset, \{\emptyset\}, \{\emptyset, \{\emptyset\}\}, \{\emptyset, \{\emptyset\}, \{\emptyset, \{\emptyset\}\}\}, \ldots$$
You will note that these are the first few ordinals that we encountered in our Axiom of Infinity, i.e.\ in von Neumann's definition of $\omega$ (see \olref[sfr][z][infinity-again]{infinity-again}). This is no coincidence. Von Neumann's definition of the ordinals treats natural numbers as ordinals, but allows for transfinite ordinals too. 

As always, we can now ask: \emph{are} these the ordinals? Or has von Neumann simply given us some sets that we can \emph{treat} as the ordinals? The kinds of discussions one might have about this question are similar to the discussions we had in \olref[sfr][rel][ref]{ref}, \olref[sfr][arith][ref]{ref}, \olref[sfr][infinite][dedekindsproof]{dedekindsproof}, and \olref[sfr][z][z-nat]{z-nat}, so I will not belabour the point. Instead, in what follows, I will simply use ``the ordinals'' to speak of ``the von Neumann ordinals''. 


\end{document}