\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}
		\olfileid{sfr}{choice}{justifications}
\olsection{Intrinsic considerations about Choice}\ollabel{justifications}
The broader question, then, is whether Well-Ordering, or Choice, or indeed the comparability of all sets as regards their size---it doesn't matter which---can be justified. 

Here is an attempted \emph{intrinsic} justification. Back in \olref[sfr][z][z-story]{z-story}, we introduced several principles about the hierarchy. One of these is worth restating:
\begin{enumerate}
	\item[] \stagesacc. For any stage $S$, and for any sets which were formed \emph{before} stage $S$: a set is formed at stage $S$ whose members are exactly those sets. Nothing else is formed at stage $S$. 
\end{enumerate}
In fact, many authors have suggested that the Axiom of Choice can be justified via (something like) this principle. I will briefly provide my own gloss on that approach.

I will start with a simple little result, which offers \emph{yet another} equivalent for Choice:
\begin{thm}[in $\ZF$]\ollabel{choiceset} Choice is equivalent to the following principle. If the !!{element}s of $A$ are disjoint and non-empty, then there is some $C$ such that $C \cap x$ is a singleton for every $x \in A$. (We call such a $C$ a {choice set} for $A$.)
\end{thm}\noindent
The proof of this result is straightforward, and I leave it as an exercise for the reader. 
\begin{prob}
	Prove \olref[sfr][choice][justifications]{choiceset}. If you struggle, you can find a proof in \cite[242--3]{Potter2004}.
\end{prob}\noindent
The essential point is that a choice set for $A$ is just the range of a choice function for $A$. So, to justify Choice, I can simply try to justify its equivalent formulation, in terms of the existence of choice sets. And I will now try to do exactly that. 

Let $A$'s !!{element}s be disjoint and non-empty. By \stageshier{} (see \olref[sfr][z][z-story]{z-story}), $A$ is formed at some stage, $S$. Note that all the !!{element}s of $\bigcup A$ are available before stage $S$. Now, by \stagesacc{}, for \emph{any} sets which were formed before $S$, a set is formed whose members are exactly those sets. Otherwise put: every \emph{possible} collections of earlier-available sets will exist at $S$. But it is certainly \emph{possible} to select objects which could be formed into a choice set for $A$; that is just some very specific subset of $\bigcup A$. So: some such choice set exists, as required.

Well, that's my best shot at a \emph{very} quick attempt to offer a justification of Choice on intrinsic grounds. But, to pursue this idea further, you should read Potter's (\citeyear[\S14.8]{Potter2004}) neat development it.

\end{document}